D3064R0: How to Avoid OOTA Without Really Trying

### 2024-03-09 ootavacuous.tex
Drop simple-reordering appendix, keep litmus test.
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Further revisions to section D.3
Here are some fairly intensive revisions to section D.3.  Excess
material is removed, the remaining parts made more concise, and the
whole thing is now consistent with the main part of the paper.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Listings~\ref{lst:OOTA With an Intra-Process Store-Load Link}
and~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized}
show the examples from
Section~\ref{sec:OOTA: rf vs. rfe}
in the form of litmus tests.
A compiler is allowed to transform the program in
Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link}
to the form of
Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized},
which demonstrates that intra-thread reads-from (rfi) links may be
eliminated by the compiler and therefore should not be used in a
definition of OOTA.
This section provides more involved examples to drive this point home.

@@

The first step towards these more involved examples is shown in
Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4},
which expands
Listing~\ref{lst:Causality Test Case 4}
(discussed in
Appendix~\ref{app:Causality Test Case 4})
from two threads to three threads.
The \co{herd7} tool finds OOTA values, as expected.
All threads have straightforward semantic dependencies from their
relaxed atomic loads to their relaxed atomic stores, which prevents this
OOTA cycle on compiler-based C++ implementations.

@@

Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
revises
Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
so that the value stored to \co{y} is forced to be at most the
value 17, and the value stored to \co{z} is forced to be at least the
value 17.
This of course means that the store to \co{z} must always be the value
17, as can be seen from the \co{herd7} output.
However, a single-thread-at-a-time analysis could not prove this.

@@

Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
further revises
Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
to flatten \co{PO()} and \co{P1()} into a single thread, so that the
resulting \co{P0()} has an rfi link instead of an rfe link connecting
the store to and load from \co{y}.
This means that a C++ compiler could prove that the value 17 is
always stored to \co{z} by examining \co{P0()} in isolation, given that
it is also legal for a C++ compiler to replace the value loaded
from \co{y} with the value just stored.\footnote{
	This is not an option in C because atomic operations are volatile,
	and thus observed behavior.}

@@

Within the confines of the C++ abstract machine (omitting the no-OOTA
restriction), this allows a value-17 OOTA cycle, courtesy of the fact
that C++ respects neither control nor data dependencies.
However, the \co{herd7} tool does not identify this cycle, perhaps due
to its restrictions on its handling of arithmetic on OOTA values.

Given a compiler that proves that the value 17 is always
stored to \co{z}, there might be no semantic dependency to
\co{P0()}'s store to \co{z}, but must be to all of the other relaxed
atomic stores.
On the other hand, if the compiler does not prove the value
stored to \co{z}, there will be a global-time delay between each of
\co{P0()}'s loads and its store to \co{z}, preventing a value-17 OOTA
cycle from forming.

Either way, the value-17 OOTA cycle cannot occur in a compiler-based
C++ implementation.
Yes, a compiler that proves that the value 17 is always
stored to \co{z} can satisfy the \co{exists} clause, but because
there is no semantic dependency to \co{P0()}'s store to \co{z},
this is not an OOTA cycle.

But what if we chose to define an OOTA cycle as (sdep $\cup$ rf) instead
of (sdep $\cup$ rfe)?

@@

In that case, it would be necessary to separately consider the dependency
from \co{P0()}'s load from \co{x} and \co{y} and that from its load
from \co{y} and its store to \co{z}.
The problem with this is shown in
Listing~\ref{lst:Why rfe Instead of Tried-And-True rf When z=17?},
which shows an equivalent program given the knowledge that the value~17
will always be stored to \co{z}.
Because all atomic operations are relaxed and because C++ does not
respect control dependencies, the store to \co{z} can be moved up
to the very beginning of \co{P0()}.
This transformation causes the rfi link associated with \co{y} to
go backwards in time, which is perfectly legitimate in this case,
but which makes defining OOTA cycles in terms of (sdep $\cup$ rf)
problematic.

Again, in C, the load from \co{y} cannot be removed because it is
volatile, and thus observable behavior.
C++ could remove that load, but we left it in to demonstrate the all-17s
outcome.

This example demonstrates why OOTA cycles must be defined in terms
of (sdep $\cup$ rfe), or in \co{herd7} nomenclature,
\co{acyclic(sdep; rfe)}, as discussed in
Section~\ref{sec:OOTA: rf vs. rfe}.

@@

\subsection{Simple Reordering as Litmus Test}
\label{app:Simple Reordering as Litmus Test}

Listing~\ref{lst:Simple Reordering as Litmus Test}
shows how
Listing~\ref{lst:Simple Reordering}
can be recast into a \co{herd7} litmus test.

>>>>

Listings~\ref{lst:OOTA With an Intra-Process Store-Load Link}
and~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized}
show the examples from
Section~\ref{sec:OOTA: rf vs. rfe}
in the form of litmus tests.
A compiler is allowed to transform the program in
Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link}
to the form of
Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized},
which demonstrates that intra-thread reads-from (rfi) links may be
eliminated by the compiler and therefore should not be used in a
definition of OOTA.
This section provides a more involved example to drive the point home.

@@

The first step in this direction is
Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4},
which expands
Listing~\ref{lst:Causality Test Case 4}
(discussed in
Appendix~\ref{app:Causality Test Case 4})
from two threads to three.
The \co{herd7} tool reports OOTA values, and
all threads have straightforward semantic dependencies from their
loads to their stores.

@@

Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
revises Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
so that the value stored to \co{y} is forced to be at most 17
and the value stored to \co{z} is forced to be at least 17.
This of course means that the value stored to \co{z} must always be
17, as can be seen in the \co{herd7} output.
(However, a single-thread analysis could not prove this fact.)

@@

Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
further revises the litmus test
to flatten \co{P0()} and \co{P1()} into a single thread, so that the
resulting \co{P0()} has an rfi link connecting
the store to and load from \co{y}, rather than the rfe link in
Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}.
As a result, even a single-thread analysis could conclude that the value
stored to \co{z} will always be 17, by examining \co{P0()} in isolation and
using the fact that it is valid for a C++ compiler to assume the value
loaded from \co{y} will always be the value just stored and so to omit
the load.\footnote{
	This is not an option in C because in that language,
	atomic accesses are volatile and hence must be preserved
	in the machine code.}

@@

If a compiler makes this observation and optimizes the program by
storing a constant 17 to \co{z} rather than going through the
computations on lines~15--17 of
Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?},
it will generate an executable which could produce the all-17's result
described by the ``exists'' clause.
(See Listing~\ref{lst:Why rfe Instead of Tried-And-True rf When z=17?},
where the store to \co{z} has been moved up to line~9.\footnote{
	The load from \co{y} on line~15 is retained even though it is
	not used, because \co{r3} appears in the ``exists'' clause and
	hence its final value is considered observable behavior.})
This would not be an OOTA cycle, because in either form of the program
the store to~\co{z} in \co{P0} is not semantically dependent on the
load from~\co{x}, so there is no cycle in (sdep $\cup$ rfe).
A different compiler that does not make this optimization will
generate an executable which cannot produce the all-17's result at all,
because of hardware dependencies and consequent instruction ordering
in the machine code.

Either way, no OOTA cycle can occur.
But what if we chose to define an OOTA cycle with (sdep $\cup$ rf) instead
of (sdep $\cup$ rfe)?

@@

In that case, it would be necessary to consider separately the dependency
in Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
connecting \co{P0()}'s load from \co{x} to its store to \co{y},
and that connecting its load from \co{y} to its store to \co{z}.
The first is a genuine semantic dependency; the second might or might
not be, depending on one's definition.
A single-thread analysis would conclude that it is, because of the
possibility that the value loaded from \co{y} might be larger than 17.
From this point of view, the all-17's result \emph{would} be
considered an OOTA cycle.

This example demonstrates another reason why defining OOTA cycles
in terms of (sdep $\cup$ rf) is problematic.
Doing so can lead to classifying a cycle as OOTA even though there
is no genuine semantic dependency in one of the participating threads.

@@

% \subsection{Simple Reordering as Litmus Test}
% \label{app:Simple Reordering as Litmus Test}

% Listing~\ref{lst:Simple Reordering as Litmus Test}
% shows how
% Listing~\ref{lst:Simple Reordering}
% can be recast into a \co{herd7} litmus test.
